---
title: "Mental Health"
author: "Maya Arnott"
date: "May 2nd, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(sf)
library(viridis)

```

# Load data.
```{r}
load("data/county_health_rankings_data.RData") 

load("data/states_boundaries.RData")
```

# Part 1 - Mapping Mental Health in the United States

## Q1: Distribution (Histogram)
```{r}

mental_health_dist <- ggplot(county_data, aes(x = poor_mental_days)) + 
  geom_histogram(bins = 30, fill = "darkblue", color = "white") +
  labs(title = "Distribution of Poor Mental Health Days",
       x = "Poor Mental Health Days",
       y = "Count")

IQR(county_data$poor_mental_days, na.rm = TRUE)

# See Q1, Median, and Q3
quantile(county_data$poor_mental_days, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)

```

Surpringly, the distribution seems to almost follow a normal distribution, with the middle 50% of the data falling between 5.25 and 6.03 average days of reported poor mental health per month. The typical mental health burden does not have a large variance, and an average of 5-6 days per month is about 18-20% of days per month, which is high.

## Q2: Choropleth Map

```{r}
chloropleth_map <- 
  ggplot() + 
  geom_sf(data = county_data, aes(fill = poor_mental_days), color = NA) +
  geom_sf(data = states_sf, fill = NA, color = "grey", size = 0.5) +
  scale_fill_viridis_c(option = "viridis", na.value = "grey50") +
  labs(title = "Poor Mental Health Days by County",
       fill = "Avg Days") +
  theme_minimal()
  
```

# Part 2 - Spatial Cluster Analysis

## Q3: Global Moran's I
